Notes

### 2022-09-02 chapters/introduction.tex
Finish document

Over the last few decades, situations involving data in the form of functions have become commonplace in many statistical scenarios, as more and more information is available worldwide with an ever-increasing level of granularity in the measurements. In particular, functional data problems are far from unheard of in the data science an machine learning community, since they have attracted the attention of researchers and practitioners equally. Medical data, weather indicators or stock exchange indices are examples of elements that benefit from a functional treatment, where the observations are regarded as single entities rather than as a conglomerate of individual points.

@@

In recent years numerous proposals have arisen on how to suitably deal with functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on data varying over a continuum. A partial survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{hsing2015theoretical} or the book by \citet{horvath2012inference}. As the name suggests, FDA techniques are heavily inspired by functional analysis tools and methods: Hilbert spaces, orthonormal systems, linear operators, and so on. In particular, a notion that also intersects with the classical theory of machine learning and pattern recognition is that of reproducing kernel Hilbert spaces (RKHS's). We will demonstrate throughout this work how these spaces of functions possess properties that allow for an efficient treatment of functional data. On the other hand, Bayesian inference methods are ubiquitous in the realm of statistics, and their usual non-parametric approach also makes use of random functions, though in a slightly different manner than in the FDA context. However, the two methodologies can certainly interact and benefit from one another, as we intend to show in this thesis. We will be particularly interested in Markov chain Monte Carlo (MCMC) methods, which allow us to approximate an arbitrary posterior distribution through a function proportional to its density.

@@

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some subtle drawbacks that emerge as a result of working in infinite dimensions. Moreover, we will concentrate our efforts on the case in which the response is a scalar, though function-on-function and function-on-scalar regression are also interesting scenarios amply explored in the literature. To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\) (i.e. a process with finite second moments and whose realizations are square-integrable functions indexed on \([0,1]\)). The underlying probability space \((\Omega, \mathcal A, \P)\) is not important. We will further suppose without loss of generality that \(X\) is centered, that is to say, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a \textit{labeled} data set \(\D_n =\{(x_i, y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), where the functional observations are recorded on a common finite grid \(\{t_j\}\subset [0, 1]\). Our ultimate aim will be to accurately predict the response corresponding to unlabeled samples from \(X\). Figure~\ref{fig:linear_data_example} depicts a typical data set used in functional regression, and we already saw in Figure~\ref{fig:tecator} what a functional classification data set may look like.

@@

It is beyond the scope of this thesis to provide a complete review of FDA as a whole, or to delve too deeply into the details and inner workings of most models and techniques mentioned. Nevertheless, references are provided frequently throughout the text to point the interested reader towards more specialized resources.

>>>>

Over the last few decades, situations involving data in the form of functions have become commonplace in many statistical scenarios, as more and more information is available worldwide with an ever-increasing level of granularity in the measurements. In particular, functional data problems are far from unheard of in the data science and machine learning community, since they have attracted the attention of researchers and practitioners equally. Medical data, weather indicators or stock exchange indices are examples of elements that benefit from a functional treatment, where the observations are regarded as single entities rather than as a conglomerate of individual points.

@@

In recent years numerous proposals have arisen on how to suitably deal with functional data, all of them encompassed under the term Functional Data Analysis (FDA), which essentially explores statistical techniques to process, model and make inference on data varying over a continuum. A partial survey on such techniques and methods is \citet{cuevas2014partial}, while a more detailed exposition of the theory and applications can be found for example in \citet{ramsay2005functional}, \citet{hsing2015theoretical} or the book by \citet{horvath2012inference}. As the name suggests, FDA techniques are heavily inspired by functional analysis tools and methods: Hilbert spaces, orthonormal systems, linear operators, and so on. In particular, a notion that also intersects with the classical theory of machine learning and pattern recognition is that of reproducing kernel Hilbert spaces (RKHS's). We will demonstrate throughout this work how these spaces of functions possess properties that allow for an efficient treatment of functional data.

On the other hand, Bayesian inference methods are ubiquitous in the realm of statistics, and their usual non-parametric approach also makes use of random functions, though in a slightly different manner than in the FDA context. However, the two methodologies can certainly interact and benefit from one another, as we intend to show in this thesis. We will be particularly interested in Markov chain Monte Carlo (MCMC) methods, which allow us to approximate an arbitrary posterior distribution through a function proportional to its density.

@@

In this work we are concerned with functional linear and logistic regression models, that is, situations where the goal is to predict a continuous or dichotomous variable from functional observations. Even though these problems can be formally stated with almost no differences from their finite-dimensional counterparts, there are some fundamental challenges as well as some subtle drawbacks that emerge as a result of working in infinite dimensions. Moreover, we will concentrate our efforts on the case in which the response is a scalar, though function-on-function and function-on-scalar regression are also interesting scenarios amply explored in the literature.

To set a common framework, throughout this work we will consider a scalar response variable \(Y\) (either continuous or binary) which has some dependence on a stochastic \(L^2\)-process \(X=X(t)=X(t, \omega)\) with trajectories in \(L^2[0, 1]\) (i.e. a process with finite second moments and whose realizations are square-integrable functions indexed on \([0,1]\)). The underlying probability space \((\Omega, \mathcal A, \P)\) is not important. We will further suppose without loss of generality that \(X\) is centered, that is to say, its mean function \(m(t)=\E[X(t)]\) vanishes for all \(t\in[0,1]\). In addition, we will tacitly assume the existence of a \textit{labeled} data set \(\D_n =\{(x_i, y_i): i=1,\dots, n\}\) of independent observations from \((X, Y)\), where the functional observations are recorded on a common finite grid \(\{t_j\}\subset [0, 1]\). Our ultimate aim will be to accurately predict the response corresponding to unlabeled samples from \(X\). Figure~\ref{fig:linear_data_example} depicts a typical data set used in functional regression, and we already saw in Figure~\ref{fig:tecator} what a functional classification data set may look like.

@@

It is beyond the scope of this thesis to provide a complete review of FDA as a whole, or to delve too deeply into the details and inner workings of most models and techniques mentioned. Nevertheless, references are often provided throughout the text to point the interested reader towards more specialized resources.

### 2022-09-02 chapters/model-choice.tex
Revision
Finish document

A well-known issue found in mixture-like models like the ones we propose is \textit{label switching}, which in short refers to the non-identifiability of the components of the model caused by their interchangeability. In our case, this happens because the likelihood is symmetric with respect to the ordering of the parameters \(b\) and \(\tau\), i.e., \(\pi(Y|X,\theta)=\pi(Y|X, \nu(\theta))\) for any permutation \(\nu\) that rearranges the indices \(j=1,\dots, p\). Thus, since the components are arbitrarily ordered, they may be inadvertedly exchanged from one iteration to the next in any MCMC algorithm. This can cause nonsensical answers when summarizing the marginal posterior distributions to perform inference, as different labelings might be mixed on each component \citep{stephens2000dealing}.

@@

The most straightforward solutions consist on imposing an artificial identifiability constraint on the parameters to break the symmetry of their posterior distributions; see \citet{jasra2005markov} and references therein. A common approach that seems to work well is to simply impose an ordering in the offending parameters, which in our case would mean requiring for example that \(\beta_i < \beta_j\) for \(i < j\), or the analogous with the \(\tau_j\). We have implemented a variation of this method described in \citet{simola2021approximate}, which works by post-processing the samples and relabeling the components to satisfy the order constraint mentioned above, choosing either \(b\) or \(\tau\) depending on which set of ordered parameters would produce the most separation between any two of them (suitably averaged across all iterations of the chains).

@@

As for the default values of the rest of hyperparameters in the prior distributions in~\eqref{eq:prior-linear}, several comments are in order:
\begin{itemize}
  \item For the expected value \(b_0\) we propose to use the MLE of \(b\). Although the likelihood function is rather involved, an approximation of the optimal value is enough for our purposes. Our numerical studies suggest that the results are much better with this choice than, say, with a random or null vector.
  \item We found that the parameter \(g\) does not have as much influence on the final result, and the experimentation indicates that \(g=5\) is a good value.
  \item Lastly, we observed that the choice of \(\eta\) can have a considerable impact on the final estimator. That is why, in an effort to normalize its scale, we consider a compound parameter \(\eta = \tilde \eta \lambda_{\max}(\mathcal X_\tau'\mathcal X_\tau)\), where \(\lambda_{\max}(\mathcal X_\tau'\mathcal X_\tau)\) is the largest eigenvalue of the matrix \(\mathcal X_\tau'\mathcal X_\tau\), and \(\tilde\eta > 0\) is the actual tuning parameter. This standardization technique has been used previously in the literature; see for example \citet{grollemund2019bayesian}.
\end{itemize}

@@

\begin{figure}[ht!]
  \centering
  \setlength{\fboxrule}{.5pt}%
  \fbox{\includegraphics[width=.65\textwidth]{ppc_linear}}
  \caption{Example of posterior predictive graphical checks on a fitted linear regression model. We show a comparison between the observed distribution of the response variable and the posterior predictive distribution consisting on the approximate sampled responses.}\label{fig:ppc_linear}
\end{figure}

In addition, we can calculate the so-called \textit{Bayesian p-values} for several statistics, which are defined as \(P(T(\symbf Y^*)\leq T(\symbf Y)| \symbf Y)\), and are computed by simply measuring the proportion of the \(M\) estimates \(T\{\symbf Y^{(m)*}\}\) that fall below the real value of the statistic. They are expected to be around 0.5 when the model accurately represents the data, and a deviation either way can be indicative of modeling issues; see Chapter 6 of \citet{gelman2013bayesian} for details.

>>>>

A well-known issue found in mixture-like models like the ones we propose is \textit{label switching}, which in short refers to the non-identifiability of the components of the model caused by their interchangeability. In our case, this happens because the likelihood is symmetric with respect to the ordering of the parameters \(b\) and \(\tau\), i.e., \(\pi(Y|X,\theta)=\pi(Y|X, \nu(\theta))\) for any permutation \(\nu\) that rearranges the indices \(j=1,\dots, p\). Thus, since the components are arbitrarily ordered, they may be inadvertently exchanged from one iteration to the next in any MCMC algorithm. This can cause nonsensical answers when summarizing the marginal posterior distributions to perform inference, as different labelings might be mixed on each component \citep{stephens2000dealing}.

@@

The most straightforward solutions consist on imposing an artificial identifiability constraint on the parameters to break the symmetry of their posterior distributions; see \citet{jasra2005markov} and references therein. A common approach that seems to work well is to simply impose an ordering in the offending parameters, which in our case would mean requiring for example that \(\beta_i < \beta_j\) for \(i < j\), or the analogous with the \(\tau_j\). We have implemented a variation of this method described in \citet{simola2021approximate}, which works by post-processing the samples and relabeling the components to satisfy the order constraint mentioned above, choosing either \(b\) or \(\tau\) depending on which set of ordered parameters would produce the largest separation between any two of them (suitably averaged across all iterations of the chains).

@@

As for the default values of the rest of hyperparameters in the prior distributions in~\eqref{eq:prior-linear}, several comments are in order:
\begin{itemize}
  \item For the expected value \(b_0\) we propose to use the MLE of \(b\). Although the likelihood function is rather involved, an approximation of the optimal value is enough for our purposes. Our numerical studies suggest that the results are much better with this choice than, say, with a random or null vector.
  \item We found that the parameter \(g\) does not have as much influence on the final result, and the experimentation indicates that \(g=5\) is a good value.
  \item Lastly, we observed that the choice of \(\eta\) can have a considerable impact on the final estimator. That is why, in an effort to normalize its scale, we consider a compound parameter \(\eta = \tilde \eta \lambda_{\max}(\mathcal X_\tau'\mathcal X_\tau)\), where \(\lambda_{\max}(\mathcal X_\tau'\mathcal X_\tau)\) is the largest eigenvalue of the matrix \(\mathcal X_\tau'\mathcal X_\tau\), and \(\tilde\eta > 0\) is the actual tuning parameter (which can be selected for example by cross-validation strategies). This standardization technique has been used previously in the literature; see for example \citet{grollemund2019bayesian}.
\end{itemize}

@@

\begin{figure}[ht!]
  \centering
  \setlength{\fboxrule}{.5pt}%
  \fbox{\includegraphics[width=.65\textwidth]{ppc_linear}}
  \caption{Example of posterior predictive graphical checks on a fitted linear regression model. We show a comparison between the observed distribution of the response variable and the posterior predictive distribution consisting on the approximate sampled responses.}\label{fig:ppc}
\end{figure}

In addition, we can calculate the so-called \textit{Bayesian p-values} for several statistics, which are defined as \(P(T(\symbf Y^*)\leq T(\symbf Y)| \symbf Y)\), and are computed by simply measuring the proportion of the \(M\) estimates \(T\{\symbf Y^{(m)*}\}\) that fall below the observed value of the statistic, \(T(\symbf{Y})\). They are expected to be around 0.5 when the model accurately represents the data, and a deviation either way can be indicative of modeling issues; see Chapter 6 of \citet{gelman2013bayesian} for details.
